\chapter{Cosmological Evolution}
\todo{Check citation -- cite vs parencite vs textcite
}

In this chapter we briefly describe the evolution of the Universe. For more comprehensive lectures see e.g. \cite{Ref:Weinberg}, \cite{2002col.luc..cosmology} or \cite{2010deto.book.....A}. We begin with basic equations for the evolution and apply them to cosmological scales. We describe different epoch in history of the Universe and also introduce number of cosmic distances often used to put observational constraints on dark energy. We show some fundamental issues with the simplest models -- horizon problem, missing (dark) matter and accelerated expansion of the Universe -- and arrive to concordance \LCDM\ model. We describe background evolution and formation of structures in \LCDM\ model and introduce some basic observables. At the end of the chapter we outline some issues with the cosmological constant and offer some basic alternative in modified gravity models.

\section{Background evolution}
\todo{cite cosmological principle}
Standard cosmological model is based on the \textit{cosmological principle}. The cosmological principle states that the Universe is homogeneous and isotropic on sufficiently large scales (beyond those traced by the large-scale structure of the distribution of galaxies). The cosmic microwave background (CMB) observations strongly support this statement as photons from different parts of the Universe are coming with almost identical temperatures. This isotropy, however, does not imply homogeneity. The \textit{Copernican Principle} states that the observer is not in a special place or time. With the observed isotropy the Copernican Principle implies the Cosmological Principle. The observed inhomogeneities and irregularities on local scales come from gravitational instability.
%%%%%%%%%%%%%%%%%%%%%%%
% Friedmann equations
%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Friedmann equations}
The homogeneous and isotropic space-time is described by Friedmann-Lema\^{i}tre-Robertson-Walker (FLRW) metric given by
\eq{
    \label{eq:flrw}
    \dd s^2=g_\uv\dd x^\mu \dd x^\nu=-\dd t^2+a^2(t)\left[\frac{\dd r^2}{1-Kr^2} + r^2\left(\dd\theta^2+\sin^2\theta\dd\phi^2 \right) \right]\,,
}
where $g_{\mu\nu}$ is a metric tensor, $a(t)$ is a scale factor with cosmic time $t$ and $K=+1,-1,0$ is a curvature that corresponds to closed, open, and flat geometries. The scale factor can be arbitrary rescaled and we adopt the common normalization $a_0=1$. We can also transform metric \eqref{eq:flrw} to a more convenient form by setting $r=\sin\chi\ (K=+1),r=\chi\ (K=0)$, and $r=\sinh\chi\ (K=-1)$. The FLRW metric is then given by
\eq{
    \label{eq:flrw_chi}
    \dd s^2=-\dd t^2+a^2(t)\left[\dd \chi^2 + f_K^2(\chi)\left(\dd\theta^2+\sin^2\theta\dd\phi^2 \right)\right]\,,
}
where
\eq{
    f_K(\chi)=
    \begin{cases}
        \sin\chi & (K=+1)\,, \\
        \chi & (K=0)\,, \\
        \sinh\chi & (K=-1)\,.
    \end{cases}
}
By allowing taking the limit $K\to0$ we can rewrite $f_K(\chi)$ in a unified way
\eq{
    \label{eq:f_K}
    f_K(\chi)=\frac{1}{\sqrt{-K}}\sinh{(\sqrt{-K}\chi)}\,.
}
In addition to the cosmic time $t$, we also introduce the conformal time $\eta$ defined by
\eq{
    \eta\equiv\int{a^{-1}\dd t}\,.
}
The dynamical equations of motion in the expanding Universe can be derived from the Einstein equations resulting in the Friedmann equations
\eq{
    \label{eq:Friedmann}
    H^2 &= \frac{8\pi G}{3}\rho - \frac{K}{a^2}+\frac{\Lambda}{3}\,, \\
    \frac{\ddot a}{a} &= -\frac{4\pi G}{3}(\rho + 3p)+\frac{\Lambda}{3}\,,\\
    \label{eq:Friedmann-continuity}
    \dot\rho &= -3H(\rho+p)\,,
}
\todo{comment $\Lambda$}

where $H\equiv\dot a/a$ is the Hubble parameter, $\Lambda$ the cosmological constant, $\rho(t)$ is energy-density and $p$ pressure. We will be also using the conformal Hubble quantity
\eq{
    \mathcal{H}\equiv\frac{1}{a}\frac{\dd a}{\dd\eta}=Ha\,.
}
These three equations are not independent and we need to also specify the equation of state, $\rho=\rho(p,t)$. The first equation is usually also written in the form
\eq{
    \Omega_m+\Omega_\gamma+\Omega_K+\Omega_\Lambda=1\,,
}
where
\eq{
    \Omega_m\equiv\frac{8\pi G\rho_m}{3H^2}\,,\hspace{3pt}
    \Omega_\gamma\equiv\frac{8\pi G\rho_\gamma}{3H^2}\,,\hspace{3pt}
    \Omega_K\equiv-\frac{K}{a^2H^2}\,,\hspace{3pt}
    \Omega_\Lambda\equiv\frac{\Lambda}{3H^2}
}

Most matter species in the Universe can be described by a simple equation of state
\eq{
    p=w\rho\,,
}
where $w$ is constant. This lead to the following solutions in flat universe
\eq{
    \rho\propto a^{-3(1+w)}\,,\hspace{6pt}a\propto (t-t_i)^{2/(3(1+w)))}\,,
}
where $t_i$ is an integration constant. Relativistic matter has the equation of state $w=1/3$ and the cosmic evolution during the radiation-dominated epoch is given by $w\propto a^{-4}$ and $a\propto(t-t_i)^{1/2}$. Non-relativistic matter has negligible pressure $(w=0)$ and the evolution during the matter- dominated era is given by $w\propto a^{-3}$ and $a\propto(t-t_i)^{2/3}$.

The accelerated expansion of the Universe $(\ddot a>0)$ is possible only for $w<-1/3$ (see Friedman equation \eqref{eq:Friedmann}), i.e. negative pressure. Note that in Newtonian gravity there is no such equivalent. The pressure is related to a force associated with a local potential that depends on the position in space but in homogeneous and isotropic universe there is no such potential.

When $w=-1$, the energy-density $\rho$ is constant and corresponds to the cosmological constant. This energy-density connected to the cosmological constant is
\eq{
    \rho_\Lambda=\frac{\Lambda}{8\pi G}\,.
}
With constant energy-density the Friedmann equations lead to an exponential expansion $a\propto\exp{Ht}$.

%%%%%%%%%%%%%%%%%%%%%%%
% Hubble`s law
%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Hubble`s law}
During the 1920s astronomers Slipher and Hubble found that the observed wavelength $\lambda_o$ of absorption lines of distant galaxies is larger than the wavelength $\lambda$ in the rest frame \cite{1929PNAS...15..168H}. This is due to the fact that the wavelength is stretched in proportion to the scale factor in an expanding Universe. The redshift $z$ is defined as
\eq{
    z\equiv\frac{\lambda_0}{\lambda} - 1=a^{-1}-1\,.
}
In an expanding Universe a physical distance $r$ from an observer (at the origin) to an object is given by $r=a(t)x$, where $x$ denotes the comoving distance. For objects which are motionless with respect to the Hubble flow the comoving distance remains constant. Taking the time derivative of $r$ we obtain
\eq{
    v_H\equiv\dot r=Hr+a\dot x
}
Because of the cosmic expansion, more distant objects are moving faster from us with velocity $v_H$. On the other hand, the peculiar velocity $v_p\equiv a\dot x$ describes the movement of an object with respect to the local Hubble flow. For small peculiar velocities and near objects $(z\ll1)$ we obtain
\eq{
    v\sim H_0r\,,
}
which is the law Hubble reported in 1929 by plotting the recessional velocity $v$ versus the distance $r$. His measurements were noisy but the current measurements give the Hubble constant the value \cite{planck_cosm}.
\eq{
    H_0=(67.4\pm0.5)\rm{\ km\ s}^{-1}\rm{Mpc}^{-1}
}

%%%%%%%%%%%%%%%%%%%%%%%
% Cosmic distances
%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cosmic distances}
In the expanding Universe there are many ways to specify the distance between two points due to the constantly changing distances and the fact that observers look back in time as they look out in distance.
\subsubsection{Comoving distance}
The light traveling along the $\chi$ direction satisfies the geodesic equation $\dd s^2=-\dd t^2+a^2(t)\dd \chi^2=0$. The light emitted at the time $t=t_1$ with $\chi=\chi_1$ (redshift $z$) reaches an observer at time$t=t_0$ with $\chi=0$ (redshift $z=0$). Integrating the geodesic equation
\eq{
    \label{eq:d_c}
    d_c\equiv\chi_1=\int_0^{\chi_1}\dd\chi=-\int_{t_0}^{t_1}\frac{\dd t}{a(t)}=\frac{1}{H_0}\int_0^z\frac{\dd\tilde z}{E(\tilde z)}\,,
}
where
\eq{
    E(z)\equiv H(z)/H_0=\sqrt{\Omega_{\gamma,0}(z+1)^4+\Omega_{m,0}(z+1)^3+\Omega_{K,0}(z+1)^2+\Omega_{\Lambda,0}}\,.
}
If we expand $E(z)$ around $z=0$ we can write the cosmic distance as
\eq{
    d_c=\frac{1}{H_0}z-\frac{E'(0)}{2H_0}z^2+\frac{2E'(0)^2-E''(0)}{6H_0}z^3+\mathcal{O}(z^4)\,,
}
where a prime represents a derivative with respect to $z$.
\subsubsection{Luminosity distance}
The luminosity distance $d_L$ is used in supernovae observations in order to link the supernova luminosity with the expansion rate of the Universe. It is defined by
\eq{
    d_L^2\equiv\frac{L_s}{4\pi F}\,,
}
where $L_s$ is the absolute bolometric (i.e., integrated over all frequencies) luminosity of a source and $F$ is the observed bolometric flux. The flux is defined by $F=L_0/S$, where $L_0$ is the observed luminosity and $S=4\pi f_K^2(\chi)$ is the area of a sphere at $z=0$.

The absolute luminosity is defined as energy emitted per unit time interval, $L=\Delta E/\Delta t$. The energy of a photon is inversely proportional to its wavelength, $E\propto\lambda^{-1}\propto 1+z$, which is stretching in an expanding universe. Also the time between arrival of two photons is proportional to the wavelength, $\Delta t\propto\lambda\propto(1+z)^{-1}$. The ratio $L_s/L_0$ is then
\eq{
    \frac{L_s}{L_0}=\frac{\Delta E_1}{\Delta E_0}\frac{\Delta t_0}{\Delta t_1}=(1+z)^2\,,
}
and the luminosity distance is
\eq{
    d_L=f_K(\chi)(1+z)\,.
}
Using definition \(f_K\) \eqref{eq:f_K} and comoving distance \eqref{eq:d_c} we can express $d_L$ as
\eq{
    \label{eq:luminosity}
    d_L=\frac{1+z}{H_0\sqrt{\Omega_{K,0}}}\sinh{\left(\sqrt{\Omega_{K,0}}\int_0^z{\frac{\dd\tilde z}{E(\tilde z)}}\right)}\,.
}
For small $z$ we can once again expand the expression and get
\eq{
    d_L=\frac{1}{H_0}z-\frac{E'(0)-2}{2H_0}z^2+\frac{2E'(0)^2-3E'(0)-E''(0)+\Omega_{K,0}}{6H_0}z^3+\mathcal{O}(z^4)\,.
}
\subsubsection{Angular diameter distance}
The angular diameter distance $d_A$ is defined as
\eq{
    d_A\equiv\frac{\Delta x}{\Delta\theta}\,,
}
where $\Delta\theta$ is the angle that subtends an object of actual size$\Delta x$ orthogonal to the line of sight. Whenever we look at objects of known size such as CMB anisotropies or BAO scale we use this distance.

The observer measures the size $\Delta x$ along the surface of a sphere with radius $\chi$ and from metric \eqref{eq:flrw_chi} follows
\eq{
    \Delta x=a(t)f_K(\chi)\Delta\theta\,.
}
The angular diameter distance is then
\eq{
    \label{eq:angular}
    d_A=a(t)f_K(\chi)=\frac{1}{1+z}\frac{1}{H_0\sqrt{\Omega_{K,0}}}\sinh{\left(\sqrt{\Omega_{K,0}}\int_0^z{\frac{\dd\tilde z}{E(\tilde z)}}\right)}\,.
}
Comparing angular diameter distance \eqref{eq:angular} and luminosity distance \eqref{eq:luminosity} we can see that they have a following relation
\eq{
    d_A=\frac{d_L}{(1+z)^2}\,.
}
\subsubsection{Degeneracy of the distance measurements}
We can se that up to the first order all the distances are the same and reduce to the Euclidean distance and that the Hubble`s law holds. With the increasing redshift the Hubble`s law dos not hold exactly and also different distances behave differently. This can be used to measure different properties of the Universe.

As the distances depend on the cosmological parameters through the integral $\int_0^z{\dd\tilde z/E(\tilde z)}$ and through $\Omega_K$ we can measure only those parameters contained in $E(z)$. Moreover, if we had distance measurements only around one particular redshift $z$ any combination of parameters that would produce similar $E(z)$ would be equally acceptable. This degeneracy can be broken by combination of measurements across different redshifts or by using other cosmological probes.

%%%%%%%%%%%%%%%%%%%%%%%
% Formation and evolution of LSS
%%%%%%%%%%%%%%%%%%%%%%%
\section{Formation and evolution of LSS}
So far we have considered only equations for smooth, homogeneous and isotropic background. In order to describe the real Universe with its rich structures we must employ perturbation theory for cosmological equations.
\subsection{Newtonian gauge}
We will consider first order (linear) perturbations of the metric:
\eq{
    g_{\mu\nu}=g_{\mu\nu}^{(0)}+\delta g_{\mu\nu}\,,
}
where $g_{\mu\nu}^{(0)}$ is the FLRW metric \eqref{eq:flrw} and all the entries in the perturbed metric $\delta g_{\mu\nu}$ have to be small with respect to the background metric. The general perturbed metric can be written as
\eq{
    \delta g_{\mu\nu}=a^2(\eta)
    \begin{pmatrix}
        -2\Psi & w_i \\
        w_i & 2\Phi\delta_{ij}+h_{ij} \\
    \end{pmatrix}\,,
}
where $\Psi(t,x)$ and $\Phi(t,x)$ are spatial scalars, $w_i(t,x)$ is 3-vector, and $h_{ij}(t,x)$ is a traceless 3-tensor. We have a freedom to choose coordinates in which we will describe the equations of motion -- we can choose our \textit{gauge}. We wish to keep our background metric $g_{\mu\nu}^{(0)}$ the same (i.e. FLRW) and only change $\delta g_{\mu\nu}$. Note that unlike ordinary coordinate transformations a gauge transformation (although expressed as a change of coordinates) does not link different observers in the same spacetime but it links two different spacetimes seen by the same observer. For a detailed discussion of gauge choices, see e.g. \cite{PhysRevD.40.1804,10.1143/PTPS.78.1,PhysRevD.22.1882}.

We will choose to attach the observers to the points in the unperturbed frame, the so called \textit{Newtonian} or \textit{longitudinal} or \textit{shear-free} gauge. In this case the observer will detect a velocity field of particles falling into the clumps of matter and will measure a gravitational potential. We can impose up to four conditions on the metric, which corresponds to the four gauge coordinate transformations. The final perturbed metric is:
\eq{
    \label{eq:flrw_pert}
    \dd s^2=a^2(\eta)\left[-(1+2\Psi)\dd\eta^2+(1+2\Phi)\delta_{ij}\dd x^i\dd x^j\right]\,.
}
\subsection{Linear perturbations}
We now decompose the Einstein tensor and the energy-momentum tensor into background and perturbed parts. The background cosmological evolution is obtained by solving the zero-th order Einstein equations and is given by the Friedmann equations. The first order equations are given by the perturbation of the Einstein tensor, i.e. by perturbed metric \eqref{eq:flrw_pert}, and by the perturbed energy-momentum tensor $T_\uv$. The energy-momentum tensor for a single perfect fluid is given by
\eq{
    T_\uv=\left(\rho+p\right)u_\mu u_\nu + pg_\uv\,,
}
where the four-velocity $u^\mu$ is up to the first order
\eq{
    u^\mu\equiv\dddd{x^\mu}{\tau}=\left[\frac{1}{a}(1-\Psi),\frac{v^i}{a} \right]\,,
}
where $\tau$ is the proper time and $v^i=\dd x^i/\dd\eta$ is the matter peculiar velocity with respect to the general expansion. We also assume that the perturbed fluid remains perfect fluid. We will use the following notation
\eq{
    \delta &\equiv\frac{\delta\rho}{\bar{\rho}}\equiv\frac{\rho-\bar{\rho}}{\bar{\rho}}\,,\\
    \theta &\equiv\nabla\cdot v\,,
}
where $\delta$ is the density contrast, the bar represents a mean value (spatial average) and $\theta$ is the velocity divergence. The first order equations are (for details, see e.g. \cite{2002col.luc..cosmology,10.1143/PTPS.78.1})
\eq{
    \label{eq:lin_1}
    3\mathcal{H}(\mathcal{H}\Psi-\Phi') + \nabla^2\Phi &= -4\pi G\bar\rho a^2 \delta\,,\\
    \label{eq:lin_2}
    \nabla^2(\Phi' - \mathcal{H}\Psi) &= 4\pi G\bar\rho a^2(1+w)\theta\,,\\
    \label{eq:lin_3}
    \Psi &=-\Phi\,,\\
    \label{eq:lin_4}
    \Phi''+2\mathcal{H}\Phi'-\mathcal{H}\Psi-(\mathcal{H}^2+2\mathcal{H}')\Psi &=-4\pi G\bar\rho a^2c^2_s\delta\,,
}
where the prime now denotes the derivative with respect to the conformal time $\eta$. Sound velocity $c_s$ is defined as
\eq{
    c_s^2\equiv\frac{\delta p}{\delta\rho}=\dddd p\rho = \frac{\dot p}{\dot\rho}\,,
}
where the last equality are valid only in the FLRW metric at background level. Conservation of the energy-momentum tensor leads to the perturbed continuity equation
\eq{
    \delta'+3\mathcal{H}(c_s^2-w)\delta=-(1+w)(\theta+3\Psi')\,,
    \label{eq:lin_5}
}
which reduces for non-relativistic matter to
\eq{
    \delta'=-\theta-3\Phi'\,.
}
Another equation coming from the conservation of the energy-momentum tensor is
\eq{
    \theta'+\left[\mathcal{H}(1-3w)+\frac{w'}{1+w}\right]\theta=-\nabla^2\left(\frac{c_s^2}{1+w}\delta + \Psi \right)\,,
    \label{eq:lin_6}
}
which reduces for non-relativistic matter to
\eq{
    \theta'+\mathcal{H}\theta=-\nabla^2\Psi-\nabla^2(c_s^2\delta)\,,
}
which is a relativistic analog of the Euler equation.


We will now transform all the equations to the Fourier space. The Fourier transformation is (up to non-relevant pre-factors)
\eq{
    f(\bm{x})=\int e^{i\bm{k\cdot x}}\hat{f}(\bm k)\dd^3k\,,
}
where $\bm k$ is a wavenumber with modulus $k$ and hat represents quantities in Fourier space. If there is no danger of misunderstanding we will drop the hat in the following. Since we are dealing with linear equations the different modes are not coupled and we can solve the equation for each $k$ independently.
 From equations \eqref{eq:lin_1}---\eqref{eq:lin_4}, \eqref{eq:lin_5} and \eqref{eq:lin_6} we will obtain the following equations in Fourier space:
\eq{
    \label{eq:lin_k_1}
    k^2\Phi + 3\mathcal{H}(\Phi'-\mathcal{H}\Psi) &= 4\pi G\bar\rho a^2 \delta\,,\\
    \label{eq:lin_k_2}
    k^2(\Phi' - \mathcal{H}\Psi) &= -4\pi G\bar\rho a^2(1+w)\theta\,,\\
    \label{eq:lin_k_3}
    \Psi &=-\Phi\,,\\
    \label{eq:lin_k_4}
    \Phi''+2\mathcal{H}\Phi'-\mathcal{H}\Psi-(\mathcal{H}^2+2\mathcal{H}')\Psi &=-4\pi G\bar\rho a^2c^2_s\delta\,,\\
    \label{eq:lin_k_5}
    \delta'+3\mathcal{H}(c_s^2-w)\delta &= -(1+w)(\theta+3\Psi')\,,\\
    \label{eq:lin_k_6}
    \theta'+\left[\mathcal{H}(1-3w)+\frac{w'}{1+w}\right]\theta &= k^2\left(\frac{c_s^2}{1+w}\delta + \Psi \right)\,,
}
where now $\theta=i\bm{k\cdot  v}$. Note that these six equations are not independent and $w$ and $c_s^2$ are arbitrary functions of time.
\subsubsection{Super-horizon scales}
In the large-scale limit $k\ll\mathcal{H}$, i.e. when the physical wavelength $\lambda_p=2\pi a/k$ of perturbations is much larger than the Hubble radius $H^{-1}$, and with the barotropic fluid, i.e. when the pressure depends only on the energy density, and with the constant $w$, we have $c_s^2=w$ and we can get an equation for $\Phi$
\eq{
    \Phi''+3\mathcal{H}(1+c_s^2)\Phi'=0\,.
}
The growing / dominant solution for $c_s^2>-1$ is $\Phi=\rm{const}$ from which follows
\eq{
    3\mathcal{H}^2\Phi=4\pi G\bar\rho a^2\delta\,.
}
Using the Friedmann equation we get $\delta=\Phi$, i.e. $\delta$ remains constant at large scales whenever $c_s^2=w$.
\subsubsection{Sub-horizon scales}
In the small-scale limit $k\gg\mathcal{H}$ we can get for a pressureless fluid $(w=0)$ with a small sound speed $c_s^2\ll1$ the Poisson equation
\eq{
    k^2\Phi=4\pi G\bar\rho a^2\delta\,,
}
and the energy conservation equation
\eq{
    \delta' &= \theta \\
    \theta' &= -\mathcal{H}\theta + c_s^2k^2\delta-k^2\Phi\,.
}
These equations lead to
\eq{
    \delta'' +\mathcal{H}\delta' +\left(c_s^2k^2 + \frac32\mathcal{H}^2 \right)\delta=0\,.
}
Perturbations on small scales with $(c_s^2k^2-\frac32\mathcal{H})>0$ do not grow but undergo damped oscillations. These are perturbations with physcial wavelength smaller than the Jeans length
\eq{
    \lambda_J=c_s\sqrt\frac{\pi}{G\bar\rho}\,.
}
For the photons and baryons before the decoupling epoch we have $c_S=c/sqrt3$ and the Jeans length is comparable to the Hubble radius $H^{-1}$ and the growth of perturbations is prevented on all scales smaller than that. For a small sound speed $c_sl\ll\mathcal{H}$ the perturbations grow freely (gravitational instability). The resulting equation for a single pressureless fluid becomes
\eq{
    \label{eq:lin_evol_m}
    \delta''+\mathcal{H}\delta'-\frac32\mathcal{H}^2\delta=0\,.
}
The growing solution during the matter dominated era is $\delta\propto a\propto t^{2/3}$ and the gravitational potential remains constant during this epoch.

%%%%%%%%%%%%%%%%%%%%%%%
% Growth function
%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Growth function and rate}
In linear perturbation theory the gravitational evolution of fluctuations is separable into time-dependent and spatially-dependent (or wavenumber-dependent) parts, i.e.
\eq{
    \delta(a)=\delta_0D(a)\,,
}
where $\delta_0$ is present overdensity and the growth function $D$ is normalized as $D(a=1)\equiv1$. The logarithmic change of growth function is the growth rate $f$
\eq{
    \label{eq:grw_rate}
    f\equiv\dddd{\ln D}{\ln a}\,.
}
 A very good approximation of a solution to density evolution \eqref{eq:lin_evol_m} is
\eq{
    f\approx|\Omega_m(a)|^\gamma\,,
}
where $\gamma\approx0.55-0.6$ depends only weakly on cosmological parameters \parencite{1980_Peebles}.

%%%%%%%%%%%%%%%%%%%%%%%
% Photon propagation
%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Photon propagation}
The above equations describe the evolution of matter density perturbations. Now we will focus on the propagation of light in this perturbed Universe. The photon momentum is defined via $k^\mu=\dd x^\mu/\dd\lambda_s$, where $\lambda_s$ is an affine parameter. The equations for the propagation are then
\eq{
    k_\mu k^\mu &= 0\,,\\
    \label{eq:p_geo}
    \dddd{k^\mu}{\lambda_s}+\Gamma^\mu_{\alpha\beta}k^\alpha k^\beta &= 0\,.
}
We will split the momentum vector into a background and a perturbed value $k^\mu=\bar k^\mu+\delta k^\mu$. The photon moves in the unperturbed metric along the direction $r$ so that $\dd\eta=\dd r$. At the background level we can integrate the photon path \eqref{eq:p_geo} to get $\bar k^0\propto a^{-2}$ and therefore the photon frequency $\nu\equiv\dd t/\dd\Lambda_s=a\bar k^0\propto a^{-1}$ as expected. The equations for the perturbed part $\delta k$ are (for details, see e.g. \cite{2010deto.book.....A})
\eq{
    \label{eq:SW_0}
    \dddd{(\delta k^0/k^0)}{\eta} &= -\left(\partpart{\Phi}{\eta}+\partpart{\Psi}{\eta}+2\partpart{\Psi}{r}\right)\,,\\
    \label{eq:WL_0}
    \frac{\dd^2x^i}{\dd\lambda_s^2}+2\mathcal{H}\dddd{\eta}{\lambda_s}\dddd{x^i}{\lambda_s} &= \left(\partpart{\Phi}{x^i}-\partpart{\Psi}{x^i}\right)\,,
}
where the two directions $x^1$ and $x^2$ are orthogonal to the propagation direction $r$.
\subsubsection{The Sachs--Wolfe effect}
The time part of the geodesic equations \eqref{eq:SW_0} describes changes in the frequency (redshift) of photons passing through changing gravitational potentials. Integrating the equation along the light-ray null path from the time of emission $e$ to the time of observation $o$ gives us
\eq{
    \left.\frac{\delta k^0}{k^0}\right\rvert_e^o=-2\left.\Psi\right\rvert_e^o-\int_{e}^{o}{\left(\partpart\Phi\eta-\partpart\Psi\eta\right)\dd\eta}\,,
}
where vertical bars represents change of quantities between $e$ and $o$. The first effect is referred to as the \textit{Sachs--Wolfe effect} and depends on the difference between the potential at emission and at observation. The second effect is called the \textit{integrated Sachs--Wolfe effect} and depends on the line-of-sight integral of $\Phi-\Psi$.
\subsubsection{Weak lensing}
The spatial part of the geodesic equations \eqref{eq:WL_0} can be written for $i=1,2$ as
\eq{
    \frac{\dd^2x^i}{\dd r^2}=\partpart{\psi}{x^i}\,,
}
where the lensing potential $\psi\equiv\Phi-\Psi$. For small displacements we can put $x^i=r\theta^i$. Integrated the above equation gives us
\eq{
    \theta^i=\theta^i_0+\int_0^r\dd r'\left(1-\frac{r'}{r}\right)\partpart{\psi(r'\theta_0^1,r'\theta_0^2,r')}{x^i}\,.
}
Two light rays separated by a small angle  $\Delta\theta^i$ on the source plane at $r=r_s$ are connected to the the observation plane at $r=0$ by the symmetric transformation matrix
\eq{
    A_{ij}\equiv\partpart{\theta^i_s}{\theta^j_o}=\delta_{ij}+D_{ij}\,,
}
where the distortion tensor
\eq{
    D_{ij}=\int_0^{r_s}\dd r'\left(1-\frac{r'}{r_s}\right)r'\frac{\partial^2\psi}{\partial x^i\partial x^j}=
    \begin{pmatrix}
        -\kappa-\gamma_1 & -\gamma_2 \\
        -\gamma_2 & -\kappa + \gamma_1
    \end{pmatrix}
    \,.
}
The convergence $\kappa$ describes the magnification of the source image and the two components of the shear field $\gamma_1,\gamma_2$ describe the distortion of the source image.
\todo{
\subsection{Non-linear cosmological perturbations}
Briefly describe 2d order, higher corrections, spherical collapse. Mention not main part -- we are interested in linear approximations.
}

%%%%%%%%%%%%%%%%%%%%%%%
% Cosmological observables
%%%%%%%%%%%%%%%%%%%%%%%
\input{chapters/cosmo_observ}

\section{Standard \LCDM model -- successes and issues}
\todo{Heyrovsky D.:

\textit{On the other hand, what is missing is an initial overview of the standard Lambda CDM model, its successes and outstanding issues (especially since the thesis focuses on developments “beyond” Lambda CDM).}
}

\todo{
\subsection{Dark matter}
Flatness and horizon problems, missing matter in clusters and galaxies
}

\subsection{Cosmic acceleration}
\todo{History, pre-1998, post 1998, $\Lambda$ and alternatives}

\subsection{Cosmological constant}

\todo{Heyrovsky D.:

\textit{
Specific comments on the scientific content: the cosmological-constant term is a natural part of the general form of Einstein’s equations, not a random ad-hoc term; most “problems” with Lambda are connected with its interpretation as vacuum energy density, they do not occur in its interpretation simply as a fundamental constant; the work could have included the observationally most relevant result: the cosmological evolution of the equation of state w(a) predicted by the different modified theories; weak lensing can be analyzed for rich galaxy clusters rather than for single galaxies.}
}

\todo{\parencite[e.g.][]{2015arXiv150205296P, 2012CRPhy..13..566M}}

\todo{Additional arguments against $\Lambda$ -- quantum corrections, etc.}
Standard cosmological \LCDM\ model is in a good agreement with all measurements of CMB \parencite{planck_cosm}, type Ia supernovae \parencite{Abbott_2019}, or BAO \parencite{BAO_results}. However, this concordance model has some significant fundamental problems and even though it may serve as an effective description of our Universe we should be looking for a deeper explanation of the accelerated expansion of the Universe.
\subsubsection{Fine-Tuning Problem}
The first problem could be phrased as the question: \textit{Why is the observed value of $\Lambda$ so small in Planck units?} In order to realize the cosmic acceleration today, we require that the cosmological constant $\Lambda$ is of the order of the square of the present Hubble parameter $H_0$. If we interpret this as an energy density, it is equivalent to
\eq{
\rho_\Lambda=\Lambda \Mpl^2\approx10^{-120} \Mpl^4\approx10^{-47}\mbox{ GeV}^4.
}
Suppose that this energy density comes from the vacuum energy. The zero-point energy of some field of mass $m$ with momentum $k$ is
\eq{
E_0=\left\langle0\left|H\right|0\right\rangle=V\int{\frac{\dd^3k}{(2\pi)^3}\frac12\sqrt{k^2+m^2}},
}
where $V=(2\pi)^3\delta^3(0)$ is the volume of space. If we trust our theory up to some cut-off scale $\Lambda_{UV}\gg m$, we obtain the vacuum energy density
\eq{
\rho_{vac}=\int_0^{\Lambda_{UV}}{\frac{4\pi k^2 \dd k}{(2\pi)^3}}\frac12\sqrt{k^2+m^2}\approx\frac{\Lambda_{UV}^4}{16\pi^2}.
}
If we take the cut-off scale to be order of the Planck scale $\Mpl\approx10^{18}$ GeV, we get a value of about 120 orders of magnitude larger than the observed value. This situation is not better for different scales in particle physics. For the SUSY\footnote{In some supersymmetric theories, the number of fermionic and bosonic degrees of freedom are equal. The energy of the vacuum fluctuations per degree of freedom is the same in magnitude but opposite in sign for fermions and bosons of the same mass. Therefore the fermion and boson contributions cancel each other and the total vacuum energy density (and consequently $\Lambda$) vanishes. As the supersymmetry has to be broken today (we do not observe supersymetry in nature) supersymmetric partners can have different masses of order $\Lambda_{SUSY}\approx10^{3}$ GeV.} scale $\Lambda_{SUSY}\approx10^{3}$ GeV, even for the QCD scale $\Lambda_{QCD}\approx0.1$ GeV, is $\rho_{vac}$ still much larger than $\rho_\Lambda$. Even if the value of $\Lambda$ does not originate from the vacuum energy there is no such a small scale in known physics -- not in GR nor in quantum physics. This huge discrepancy between theoretical predictions and observed value of $\Lambda$ is a major issue in physics and cosmology.
\subsubsection{Coincidence Problem}
The second problem  could be phrased as the question: \textit{Why is the energy density $\rho_\Lambda$ so close to the present matter density?} While the energy density of the cosmological constant (vacuum energy) remains constant in time and was completely negligible in most of the past, the energy density of matter evolves like $\rho_m\propto(1+z)^3$ and will be entirely negligible in the future. According to the Copernican principle \parencite{ellis_maartens_maccallum_2012} we do not live in a special place \textbf{nor time}. Thus it is very unlikely that these two components will have densities of the same order of magnitude in the present. If $\rho^{(0)}_\Lambda/\rho^{(0)}_m$ was just 10 or 100 times smaller, we would not see any accelerated expansion. If it were a few orders of magnitude larger than one, the transition to the accelerated universe would occur at a large redshift.
\\

The so-called anthropic principle can seemingly give us the explanation for both of the two cosmological constant problems, why it is small and why the acceleration starts now. Because if the vacuum energy has been big and dominant from the earlier epoch, there would be no chance to form structures in the Universe, like galaxies, stars, planets and us, intelligent lives. But this anthropic explanation of the value $\rho_\Lambda$ makes sense only if there is a multiverse with a lot different realizations of $\rho_\Lambda$. But we live only in our realization of the Universe and therefore we cannot verify whether the anthropic principle solves the cosmological constant problem or not.

Beside these two problems concerning the cosmological constant, the observations of coherent acoustic oscillations in the CMB \parencite{planck_cosm} has turned the notion of accelerated expansion in the very early universe (inflation) into an integral part of the cosmic standard model. This early accelerated expansion was not due to the cosmological constant, because in that case the inflation would not stop and today`s universe would not be possible. Therefore, we have to postulate some new scalar field (inflaton) \parencite{2015IJMPD..2430010C} that we know so little about. If we do not properly understand the past dynamics of the universe how can we accept the cosmological constant as an explanation of the present acceleration without doubt?