\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}
% Do autoref \url{http://home.ef.jcu.cz/~houda/publications/2009-dissertation-abstract.pdf}
The~thesis was dealing with~the~topic of~modified gravity and~how we can discover its resulting deviations from general relativity. In~\autoref{chpt:cosmo_evol} we summarized linear equations for~the~evolution of~the~Universe. We described how we can use the~evolution process to~studies of~cosmological parameters of~the~Universe. We described basic cosmological observables and~their usefulness when distinguishing between different cosmologies.

\hyperref[chpt:de_mg]{Chapter 2} was focused on~modifications of~the~general relativity. We described successes and~failures of~the~standard cosmological model and~presented main issues with~the~cosmological constant. We described how we can modify the~theory of~gravity and~focused on~one of~the~most studied extensions -- the~\fR\ gravity. We studied the~Hu-Sawicki \fR\ theory from a~different angle of~view than most authors -- as the~chameleon gravity in~the~Einstein frame. We studied the~chameleon behavior numerically and~we checked the~linear predictions with~our numerical solutions. We studied the~chameleon behavior in~spherical systems (stars, galaxies, clusters) and~concluded that the~chameleon mechanism hides the~fifth force on~scales smaller than superclusters for~the~Hu-Sawicki model and~that the~chameleon needs to~be studied on~large cosmological scales through \nbodysim s. Our original method of~solving modified gravity in~spherical systems represents an~important contribution because it can be applied to~other models than Hu-Sawicki \fR. With~these methods we can check the~behavior of~other theories in~spherical systems and~also check their behavior in~laboratory conditions on~the~Earth.

In~\autoref{chpt:cosmo_sim} we described general techniques when dealing with~cosmological \nbodysim s and~how we implemented them in~our own code for~\nbodysim s -- with~both standard and~modified gravity. We described our own contribution to~the~\code{CCL} code and~how the~\code{CCL} can help scientists worldwide to~speed up their work on~cosmology.

In~\autoref{chpt:app_schemes} we introduced different approximations that can be used to~study the~Universe quickly and~intuitively. Most of~the~previous studies focused on the~Einstein--de Sitter Universe whereas in~this work we generalized the~equations to~the~\LCDM\ cosmology. We described how we implemented these approximations in~studies of~modified gravity which has not been done before and~is one of~the~original result of~this work.

Key results of~the~thesis are presented in~\autoref{chpt:app_sims}. We described original results of~our cosmological \nbodysim s using approximate schemes. We focused on~aspects that have drawn less attention in~the~past -- slower growth of~structure formation (modified growth function), study of~the~correlation function, predictions of~the~shape of~the~peak of~baryonic acoustic oscillations in~real space and~the~ability to~predict certain non-linear features of~full \nbodysim s. We showed that these approximations can be used to~study scales around the~baryon acoustic oscillation scale, $k\sim 0.1~h\text{Mpc}^{-1}$ but not much further. We also tested these methods on~the~chameleon gravity and~showed what probes are suited the~best for~distinguishing between different parameters of~the~chameleon gravity. Unlike matter power spectra and~baryonic acoustic oscillation, the~halo mass function does not seem to~be a~good way to~study the~chameleon gravity. In~\autoref{chpt:outlook} we then discussed other possible applications of~approximate methods described in~this work and~we outlined further course of~our studies.

In~\autoref{chpt:cosmo_surveys} we reviewed the~present-day and~future experiments designed to~study our Universe. We showed what constraints on~our Universe have already been observed and~what accuracy of~future experiments we can expect.

\subsubsection{Author's publications}
\begin{refsection}[bibliography/my_work.bib]
In~the~\hyperref[chpt:list_publish]{List of~publications} all author's publications are listed (in~alphabetical order by the~authors' last names). The~list includes the~author's previous work in~the~Cherenkov Telescope Array collaboration \parencite{2016arXiv161005151C,2017arXiv170903483A,2017ApJ...840...74A,2019scta.book.....C}. While working on~the~CTA, the~author's main interest were the~atmospheric simulations of~cosmic showers \parencite{2017EPJWC.14401014V,}. The~purpose of~these simulations is to~improve the~calibration of~the~atmospheric properties as well as a~calibration of~the~detector response. One of~the~main contributions to~the~systematic uncertainties of~the~CTA measurements stems from the~uncertainty on~the~atmospheric density profile, of~molecules and~aerosols, which these simulations help to~reduce.

The~author's work inside the~Dark Energy Science Collaboration (DESC) focused mainly on~improving the~Core Cosmology Library \parencite[\code{CCL},][]{2019ascl.soft01003C,2019ApJS..242....2C}. Initially, the~author helped to~create a~documentation of~the~library, wrote many examples of~usage of~the~\code{CCL}, and~presented with~other authors the~\code{CCL} and~its capabilities at~several sessions to~new users. The~author also helped to~improve several automatization processes regarding the~releasing of~the~library and~helped to~improve its compatibility across different operating systems and~environments.

The~author's main work, \textit{Fast approximate methods for~modified gravity cosmological simulations} \parencite[published in~Monthly Notices of~the~Royal Astronomical Society,][]{2020MNRAS.493.2085V}, summarizes results regarding the~cosmological \nbodysim s described in~this work (\autoref{chpt:cosmo_sim} -- \autoref{chpt:app_sims}). It is the~result of~several-year research regarding the~cosmological simulations. During this time, the~author wrote his own code for~cosmological simulations, implemented methods for~solving highly non-linear equations, and~developed complex pipelines for~processing and~analyzing data of~these simulations.
\end{refsection}
