\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}
Since the twentieth century, astronomers have accumulated conclusive evidence that the content of the Universe is mostly of unknown origin and that the ordinary matter, baryonic matter, constitutes a tiny fraction of the energy density of the Universe -- only 5\%. The remaining bulk of the Universe is composed of 70\% of dark energy causing the accelerated expansion of the Universe, whereas 25\% is in a form of a dark matter causing the formation of the structures in the Universe. Dark energy ranks as one of the most important discoveries in cosmology, with profound implications for astronomy and fundamental physics.

With the measurement of an accelerated rate of expansion of the Universe, various alternatives to standard Einstein's theory of gravity have been suggested to explain the acceleration, attempting to bypass problems connected to the cosmological constant. These theories usually add some new degrees of freedom, either in a form of new fields or by modifying existing fields.

There are many different ways how to study the dark energy (or modified gravity), e.g. through a growth of structures on large scales and consequent formation of cluster and galaxies on smaller scales, bending of light in the Universe, or baryonic oscillations in the early Universe.

Predictions of structure formation and other observables cannot be obtained analytically even for the standard theory of gravity, let alone for highly non--linear equations of modified gravity. One must employ numerical methods such as \nbodysim s or perturbative methods. What is possible to study numerically in the fully non--linear regime for standard gravity can be very expensive for modified gravity, and it is difficult, if not impossible, to explore such a huge space of models and parameters using high-resolution \nbodysim s. Even in the mildly non--linear regime, perturbative methods can become extremely complex.

Due to the numerical difficulties connected with highly non--linear equations of modified gravities, one must often look for new ways how to address these problems. We explore whether simplified dynamical approximations, applicable for a certain set of cosmological probes, can be used to investigate models of modified gravity with acceptable accuracy in the latter instance.

For our purposes, from many different models of modified gravity, we chose to explore and test these simplified methods on the Hu-Sawicki \fR\ model. \fR\ gravity represents a broad class of theories. It can be considered as the simplest example of an extended theory of gravity. \fR\ gravity can be also studied from a different point of view than a modification of gravity. We can see this extra degree of freedom as a new scalar field, chameleon field, with strong non-standard gravitation coupling to other fields. The chameleon field has a mass--dependent on surrounding density and can, therefore, escape standard Solar system tests through this so-called chameleon mechanism. We will study this mechanism on galaxy scales, cluster scales, and mainly on large cosmological scales.

All theoretical predictions which can be obtained analytically or numerically through simulations need to verify by real--life experiments. Present-day experiments such as DES, BOSS, or Planck place constraints on modified gravities and so far no extension of Einstein`s gravity has been confirmed but many alternatives remain viable. But a new era of next--generation experiments is coming in the next decade, such as the Vera C. Rubin Observatory, Euclid, or W-FIRST.

The Vera C. Rubin Observatory, previously referred to as the Large Synoptic Survey Telescope (LSST), is a ground-based telescope being built in northern Chile. Thanks to a large aperture, wide--field survey telescope, and 3200 Megapixel camera it will image faint astronomical objects across the sky. The LSST will rapidly scan the sky, charting objects that change or move: from exploding supernovae to potentially hazardous near-Earth asteroids. The LSST will produce a very deep survey and its images will trace billions of remote galaxies, providing multiple probes of the dark matter and dark energy.

The thesis is organized as follows: in \autoref{chpt:cosmo_evol} we review basics of the evolution of the Universe; in \autoref{chpt:de_mg} we describe various theories of dark energy and modified gravity while focusing on the chameleon theory. We study the non--linear behavior of the chameleon field numerically in systems exhibiting spherical symmetry in \autoref{sec_cham}, which is something not studied previously. In \autoref{chpt:cosmo_sim} we describe different techniques when dealing with large cosmological simulations, how we adapted them in our own code for \nbodysim s, and our contributions to a publicly available software \code{CCL} which compute basic cosmological observables. In \autoref{chpt:app_schemes} we introduce different approximations, those we use in our simulations, and also approximations that are used by other codes or have been studied in the past. Unlike previously studied cases of Einstein--de Sitter universes we adapted these approximations to general \LCDM\ cosmologies. In \autoref{chpt:app_sims} we describe original results of our cosmological \nbodysim s using approximate schemes. In this chapter, we also study the chameleon gravity behavior on cosmological scales. In \autoref{chpt:outlook} we discuss possible future applications of implemented techniques and ways to further improve them. In the end, in \autoref{chpt:cosmo_surveys} we review present-day and future experiments designed to study our Universe.

\section*{Units and conventions}
Throughout this work, we use units such that $c=\hbar=k_B=1$, where $c$ is the speed of light, $\hbar$ is reduced Planck`s constant, and $k_B$ is Boltzmann`s constant. We list frequently used symbols after the Table of Contents. We adopt the metric signature $(-, +, +, +)$. The Greek indices such as $\mu$ and $\nu$ run from 0 to 3 whereas the Latin indices such as $i$ and $j$ run from 1 to 3. When referring to present values of quantities we use subscript $0$, e.g. $H_0$ to denote the present value of the Hubble parameter.